Novel Methods for Characterizing Peaks from Direct Injection FT-MS Experiments

### 2022-03-10 doc/peakcharacterization_manuscript.Rmd
trying to get most of the functionality there!

---
title: "Novel Methods for Characterizing Peaks from Direct Injection FT-MS Experiments"
author: "Robert M Flight, Joshua M Mitchell, Hunter NB Moseley"
date: "`r Sys.time()`"
output: 
  redoc::redoc
---

```{r setup, echo = FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, dpi = 600, dev = "png", dev.args = list(type = "cairo"))
```

@@

To find intervals that contain points that are not just noise, the number of non-zero intensity frequency points in each sliding window are counted.
The 99th percentile of non-zero counts across 10000 sliding windows are computed over all sliding windows, and then the median value of these 99th percentiles computed. 
Any sliding window with a non-zero count less than or equal to 1.5 times the median percentile is removed, and the remaining sliding windows are reduced, where overlapping windows are merged together to create the initial peak regions.

@@

```{r mz_frequency_figure}
drake::loadd(frequency_conversion)
frequency_conversion
```

*caption*: **A**: Intensity vs M/Z for a single peak from a single scan. 
Red lines denote the differences between each point, and red dots the average between the pair of points. 
The difference over the point is used to derive the frequency values in **B**. 
**B** plots the intensity vs the converted frequency points derived from **A**. 
The red lines denote pairwise differences, which are shown in **C** for this single peak. 
The differences for all peaks across all scans vs M/Z are shown in **D**, with those differences that lie within 0.49 - 0.51 shown in red. 
**E** shows the plot of derived frequency *vs* M/Z, with fitted values from the linear model in red.

@@

```{r peak_ordering}
drake::loadd(peak_ordering)
peak_ordering
```

*caption*: Peak ordering in M/Z compared with ordering in frequency space when a single M/Z to frequency model is used or scan specific M/Z to frequency models are used. 
For a single peak, the scan level peak M/Z's were extracted, and then frequency values for those M/Z generated using a single common model of M/Z to frequency (*single_order*), or models derived from each scan (*scan_order*). 
A subset of the peaks end up out of order using scan specific models, implying that a single model should be used across all the scan level data.

@@

```{r slidingwindow_count_distribution}
drake::loadd(sliding_regions)
sliding_regions
```

*caption*: Histogram of the number of non-zero points across all scans in sliding windows 10 frequency points wide and 1 frequency point apart. 
The red vertical line denotes the lower 99th percentile of the data, only sliding windows with non-zero counts above the red line will be kept for subsequent use.

@@

```{r centroided_peaks}
# figure showing fit of quadratic model to actual data points
drake::loadd(peak_fit_plot)
peak_fit_plot
```

@@

With the characterized (centroided) peak data from across scans within each region, it is then important to determine if only one or multiple "peaks" are actually present in the region. 
Our solution to this is to define breaks between **actual** peaks as a single frequency bin with zero characterized peaks within it. 
The frequency bins are created from tiled windows that are one frequency point difference wide. 
Adjacent non-zero frequency bins are merged to comprise a single peak region. 
Figure X shows an example where an initial region is broken up into two regions based on the characterized peak centers.

```{r breaking_regions}
# figure showing multiple "peaks" in what is initially a single region
drake::loadd(split_region_plot)
split_region_plot
```

*caption*:Splitting a single region into two regions based on the present peaks.
**A**: The full set of raw frequency and intensity data across all scans for the region are shown. 
Clearly the region has two separate peaks within it.
**B**: The peak centroids (frequency and intensity) for each peak in black.
The tiled regions (red) are used to quantify the number of peaks.
**C**: The number of peaks within each tiled region shown as a histogram.
Each group of non-zero adjacent regions will be merged to form a new peak region.

@@

```{r normalization_data}
# figure showing peak intensity correlation with M/Z, intensity varying with M/Z
drake::loadd(correlate_scan_height_plot_correlate_scan_height_data)
correlate_scan_height_plot_correlate_scan_height_data
```

@@

```{r normalization_factors}
drake::loadd(normalization_plot_normalization_values_data)
normalization_plot_normalization_values_data
```

@@

@@

```{r rsd_plots}
drake::loadd(rsd_plot_rsd_data)
rsd_plot_rsd_data
```

*caption*: Density plots of relative standard deviations (RSD) of peak heights across scans for each of the processing methods. 
A peak had to be present in at least three scans for the RSD value to be reported.

```{r rsd_table, results = 'asis'}
drake::loadd(rsd_table_rsd_data)
knitr::kable(rsd_table_rsd_data, digits = 3)
```

>>>>

---
title: "Novel Methods for Characterizing Peaks from Direct Injection FT-MS Experiments"
author: "Robert M Flight, Joshua M Mitchell, Hunter NB Moseley"
date: "`r Sys.time()`"
output: 
  word_document:
    keep_md: true
editor_options: 
  chunk_output_type: console
---

```{r setup, echo = FALSE}
knitr::opts_chunk$set(echo = FALSE, 
                      warning = FALSE, 
                      message = FALSE, 
                      dpi = 600, 
                      dev = "png", 
                      dev.args = list(type = "cairo"),
                      fig.process = dn_figure_rename)
```

```{r get_setup}
source("./packages.R")
lapply(list.files("./R", full.names = TRUE), source)

figure_count = dn_counter$new("Figure ")
```

@@

To find intervals that contain points that are not just noise, the number of non-zero intensity frequency points in each sliding window are counted.
Subsequently, we break these counts into fixed width tiles (default width of 2000) and calculate the 99th percentile of non-zero points for each tile.
The median value x 1.5 of these 99th percentile values from the fixed width tiles is used as the cutoff value to determine which of the initial sliding regions should be kept as regions that contain potential signal.
Any sliding window with a non-zero count less than or equal to the cutoff value is removed, and the remaining sliding windows are reduced, where overlapping windows are merged together to create the initial peak regions.

**Hunter talk about issues around non-detection and why we only use non-zero counts, sliding followed by tiles**.

@@

```{r increment_mz_frequency}
figure_count$increment("mz_frequency_conversion")
```

```{r mz_frequency_conversion, dn_id = figure_count}
tar_load(frequency_conversion)
frequency_conversion$all
```

`r figure_count$label_text("mz_frequency_conversion")`. **A**: Intensity vs M/Z for a single peak from a single scan. 
Red lines denote the differences between each point, and red dots the average between the pair of points. 
The difference over the point is used to derive the frequency values in **B**. 
**B** plots the intensity vs the converted frequency points derived from **A**. 
The red lines denote pairwise differences, which are shown in **C** for this single peak. 
The differences for all peaks across all scans vs M/Z are shown in **D**, with those differences that lie within 0.49 - 0.51 shown in red. 
**E** shows the plot of derived frequency *vs* M/Z, with fitted values from the linear model in red.

@@

```{r increment_peak_ordering}
figure_count$increment("peak_ordering")
```

```{r peak_ordering, dn_id = figure_count}
tar_load(peak_ordering)
peak_ordering
```

`r figure_count$label_text("peak_ordering")`. Peak ordering in M/Z compared with ordering in frequency space when a single M/Z to frequency model is used or scan specific M/Z to frequency models are used. 
For a single peak, the scan level peak M/Z's were extracted, and then frequency values for those M/Z generated using a single common model of M/Z to frequency (*single_order*), or models derived from each scan (*scan_order*). 
A subset of the peaks end up out of order using scan specific models, implying that a single model should be used across all the scan level data.

@@

```{r increment_slidingwindow}
figure_count$increment("slidingwindow_count")
```

```{r slidingwindow_count, dn_id = figure_count}
tar_load(sliding_regions)
sliding_regions
```

`r figure_count$label_text("slidingwindow_count")`. Histogram of the number of non-zero points across all scans in sliding windows 10 frequency points wide and 1 frequency point apart. 
The red vertical line denotes the lower 99th percentile of the data, only sliding windows with non-zero counts above the red line will be kept for subsequent use.

@@

```{r increment_fit}
figure_count$increment("centroided_peaks")
```

```{r centroided_peaks, dn_id = figure_count}
# figure showing fit of quadratic model to actual data points
tar_load(peak_fit_plot)
peak_fit_plot
```

@@

```{r increment_breaking_regions}
figure_count$increment("breaking_regions")
```

With the characterized (centroided) peak data from across scans within each region, it is then important to determine if only one or multiple "peaks" are actually present in the region. 
Our solution to this is to define breaks between **actual** peaks as a single frequency bin with zero characterized peaks within it. 
The frequency bins are created from tiled windows that are one frequency point difference wide. 
Adjacent non-zero frequency bins are merged to comprise a single peak region. 
`r figure_count$label_text("breaking_regions")` shows an example where an initial region is broken up into two regions based on the characterized peak centers.

```{r breaking_regions, dn_id = figure_count}
# figure showing multiple "peaks" in what is initially a single region
tar_load(split_region_plot)
split_region_plot
```

`r figure_count$label_text("breaking_regions")`. 
Splitting a single region into two regions based on the present peaks.
**A**: The full set of raw frequency and intensity data across all scans for the region are shown. 
Clearly the region has two separate peaks within it.
**B**: The peak centroids (frequency and intensity) for each peak in black.
The tiled regions (red) are used to quantify the number of peaks.
**C**: The number of peaks within each tiled region shown as a histogram.
Each group of non-zero adjacent regions will be merged to form a new peak region.

@@

```{r normalization_data, eval = FALSE}
# figure showing peak intensity correlation with M/Z, intensity varying with M/Z
tar_load(correlate_scan_height_plot_correlate_scan_height_data)
correlate_scan_height_plot_correlate_scan_height_data
```

@@

```{r normalization_factors, eval = FALSE}
tar_load(normalization_plot_normalization_values_data)
normalization_plot_normalization_values_data
```

@@

```{r increment_rsd}
figure_count$increment("rsd_method")
```

@@

```{r rsd_method, dn_id = figure_count}
tar_load(rsd_plot)
rsd_plot
```

`r figure_count$label_text("rsd_method")`. 
Density plots of relative standard deviations (RSD) of peak heights across scans for each of the processing methods. 
A peak had to be present in at least three scans for the RSD value to be reported.

```{r rsd_table, results = 'asis'}
tar_load(rsd_values)
knitr::kable(rsd_table_rsd_data, digits = 3)
```
